\documentstyle[a4wide,11pt,parskip]{ukart}
\begin{document}
\pagestyle{empty}
\begin{center}\LARGE \bf Laser-Scan Ltd\end{center}

\begin{center}\LARGE \bf SPR Response\end{center}

\vspace{10mm}
Customer: Nottingham University\\
Your ref: Gary Priestnall/10/10/92\\
Our ref: SPR 1988 logged 15/10/92\\
Date: 18th March 1993\\
From: Paul Hardy \& John Barber.

\section{Introduction}
This response applies to an SPR submitted about failures in polygon forming,
involving the ILINK and IPOLYGON programs.
 
We are deeply sorry about the delay in responding to this SPR. This is due to a 
combination of several unfortunate events:
\begin{itemize}
\item
When the supporting data for your SPR arrived in October, we did carry out our
usual immediate technical evaluation. Unfortunately, your problem, although
very well described in your SPR, was not simply reproducible here.
When the procedure you supplied was run, the programs appeared to work
satisfactorily, producing the requisite 3 polygons. The technical reasons for
this are explained below. However, due to this, the SPR was classified in
November as ``non-trivial'' and put on a waiting list for resource allocation
for further  investigation.
\item
There was a weakness in the procedures which should have
been escalating the urgency and current state of your SPR. This breakdown has
been recognised, and addressed, in that regular SPR review meetings have
recently restarted, under the control of our new Quality Manager (Andy
Giddings). Your SPR was in a group which have recently been picked out as not
having received adequate attention, and are now being addressed. 
\item
Because your SPR stated that the the problem only
arose after the software upgrade to HORIZON V1.3, it had been assumed (incorrectly)
that you would have reverted to the versions of ILINK and IPOLYGON from before
the upgrade, as a workround.
\item
Between November 1992 and February 1993 there has been extensive work done on
ILINK to correct some deep-rooted problems with certain fundamental algorithms.
Work on various associated SPRs was suspended until this algorithm work was
complete, as it was felt that some of the problems described by these SPRs
would go away as a side-effect of the algorithm changes. The algorithm changes
have been included in ILINK V5.0, which is currently shipping as part of the
March 1993 LAMPS V4.1 release. 

ILINK V5.0 does indeed fix the majority of outstanding SPR problems. The effect
of this bulge of work, however, was that a small number of outstanding ILINK
problem SPRs were not able to be addressed in the V5.0 release.  However the
software engineer involved is continuing his efforts to address them all, and
your SPR would have been tackled next.
\end{itemize}

Following Roy Haines-Young's letter of 17th March 1993, it has been escalated to the top of the list and
additional senior 
technical staff have also become involved. The cause has now been isolated, 
workrounds are presented, and a complete solution is under way.

\section{Problem restatement}
\begin{itemize}
\item
A flowline is presented which uses ILINK/STRUCTURE to form link/node geometry,
followed by IPOLYGON to form complete nested polygons.
\item
IPOLYGON reports ``Unable to find current arm in feature with FSN xxx --  Polygon
abandoned''.
\item
The output polygon files contain missing polygons.
\end{itemize}

\section {Problem Analysis}
\begin{itemize}
\item
Initial analysis was hindered because the problem was not reproducible.
When the procedure you supplied was run, the programs appeared to work
satisfactorily, producing the requisite 3 polygons. Further analysis was
suspended awaiting scheduling resources.
\item
However, when the SPR was re-addressed later, the same procedure when run using
the same versions of the same programs and data files reproduced the error!
\item
The IFF data files were compared carefully, and it was noticed that the
successful run had used 2-dimensional IFF coordinate string entries (ST), while
the unsuccessful run had used multi-dimensional IFF coordinate block (CB)
entries (actually still just containing 2 dimensions -- X and Y).
\item
The junction pointer structure of the CB version was then analysed, and found
that one arm of one junction pointed 4 bytes further on in the file than it
should.
\item
All other arms and junctions were correct, but the offending arm was the only
one which pointed at the end of a feature made up of more than one coordinate
block entry (more than 200 points).
\item
This was traced through the program and tracked down to ILINK assuming that it
could determine the offset of the final coordinate block based on the number of
points. This was true for 2-dimensional ST entries, but not for the
multi-dimensional CB entries.
\item
The incorrect pointer was produced only when three conditions were satisfied:
\begin{itemize}
\item
IFF Revision level 1 was active, so that CBs were used.
\item
The feature has more than 200 points.
\item
The junction is at the end not the start of the feature.
\end{itemize}
\item
A fix to the problem will involve changing the program to remember the correct
pointer as the CB entries are output, rather than trying to calculate it later.
This fix is not trivial, but well defined.
\end{itemize}

\section {Workrounds and Solution}
\begin{itemize}
\item
As a workround, deassign the logical name ``LSL\$IFF\_OUTPUT\_REVISION before
running ILINK. This will cause only 2 dimensional coordinates to be output, and
avoid the problem.
\item
An alternative workround is to divide the input IFF data so that no feature is
more than 200 points long. The ICUT utility of the IMP package can be used to
do this.
\item
A full fix for this problem is under way at present. We will make it available
to you as soon as the version of ILINK containing the fix reaches Beta test
status.
\end{itemize}

\section{Conclusions}
\begin{itemize}
\item
The reported problem is caused by ILINK writing bad junction pointer information
into the structured IFF file. 
\item
The effect is due to a coding fault made in 1988 when ILINK was enhanced to
support multi-dimensional coordinates and per-point attributes (CBs).
\item
The problem does not occur when the programs are used in the default
traditional XY coordinate string IFF environment. It only occurs when the ``IFF
Revision Level 1'' facilities are activated so as to support per-point
attributes. 
\item
Reverting to IFF revision level 0 may well be an adequate workround 
for this problem in the short term.
A full fix for this problem is under way at present. 
\item
This SPR did not get adequate response or escalation, due to a combination of
effects. Laser-Scan apologise again for this, and have already taken steps to
ensure that similar situations do not occur again.
\item
If you want to discuss the technical basis for this SPR response, feel free to 
telephone me on 0223 420414.
\end{itemize}
\vspace{10mm}
\center{\it Paul Hardy\\
Chief Programmer}
\end{document}
